---
output:
  md_document:
    variant: markdown_github
  html_document:
    variant: markdown_github
    keep_md: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center",
  fig.retina = 2,
  out.width = "75%",
  dpi = 96
)
knit_hooks$set(pngquant = hook_pngquant)
```

# intradayModel
Our package uses state-of-the-art state-space models to facilitate the modeling and forecasting of financial intraday signals. This package currently offers a univariate model for intraday trading volume, with new features on intraday volatility and multivariate models in development. It is a valuable tool for anyone interested in exploring intraday, algorithmic, and high-frequency trading.

## Installation
The package can be installed from [GitHub](https://github.com/convexfi/intradayModel):
```{r, eval=FALSE}
# install development version from GitHub
devtools::install_github("convexfi/intradayModel")
```

Please cite **intradayModel** in publications:
```{r, eval=FALSE}
citation("intradayModel")
```

## Quick Start
To get started, we load our package and some sample data: the 15-minute intraday trading volume of AAPL from 2019-01-02 to 2019-06-28, covering 124 trading days.

```{r, include=FALSE}
devtools::load_all(".")
```

```{r, message = FALSE}
library(intradayModel)
data(aapl_volume)
aapl_volume[1:5, 1:5] # print the head of data
```

Next, we fit a univariate state-space model using `fit_unimodel` function. To be specific, we use the first 104 trading days for fitting, and the last 20 days for evaluation of forecasting performance.

```{r}
aapl_volume_fit <- aapl_volume[, 1:104]
unimodel_obj <- fit_unimodel(aapl_volume_fit)
```

Once the model is fitted, we can estimate the hidden components of any intraday signal based on all its observations, which is called **smoothing**. By calling `smooth_unimodel` function, we obtain daily, seasonal, and intraday dynamic components. This procedure helps us better identify the underlying information of the intraday signal.

```{r smooth_plot}
smooth_result <- smooth_unimodel(aapl_volume_fit, unimodel_obj)
plot_components(smooth_result) # plot smoothed hidden components
plot_performance(smooth_result) # plot smoothed result
```

To see how well our model performs on new data, we use `forecast_unimodel` function to do one-bin-ahead forecast on the out-of-sample dataset of 20 days.

```{r forecast_plot}
forecast_result <- forecast_unimodel(aapl_volume, unimodel_obj, out.sample = 20)
plot_components(forecast_result) # plot forecast hidden components
plot_performance(forecast_result) # plot forecast result
```

## Contributing
We welcome all sorts of contributions. Please feel free to open an issue
to report a bug or discuss a feature request.

## Citation
If you made use of this software please consider citing:

- Chen, R., Feng, Y., and Palomar, D. (2016). Forecasting intraday trading volume: A kalman filter approach. <https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3101695>

## Links
Package: [GitHub](https://github.com/convexfi/intradayModel)

README file: [GitHub-readme](https://github.com/convexfi/intradayModel/blob/master/README.md).

Vignette: [GitHub-vignette](https://htmlpreview.github.io/?https://github.com/convexfi/intradayModel/blob/master/vignettes/intradayModel.html).
