---
title: "Intraday Volume and Volatility: Modeling and Prediction"
author: |
  | Shengjie Xiu, Yifan Yu, and Daniel P. Palomar
  | The Hong Kong University of Science and Technology (HKUST)
date: "`r Sys.Date()`"
output: 
  cleanrmd::html_document_clean:
    theme: bamboo
    toc: yes
    toc_depth: 2
csl: apalike.csl # springer-socpsych-author-date.csl
csl-entry-spacing: 1.2em
bibliography: refs.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Intraday Volume and Volatility: Modeling and Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteKeyword{State-space Model}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.retina = 2,
  out.width = "85%",
  dpi = 96,
  pngquant = "--speed=1"
)
knit_hooks$set(pngquant = hook_pngquant)  # brew install pngquant
options(width=100)
```

```{css, echo = FALSE}
/* ensure cleanrmd is centered */
body {
  margin: 0 auto;
  max-width: 750px;
  padding: 2rem;
}
```

-----------
> This vignette illustrates the usage of the package `intradayModel` for modeling and prediction of intraday trading volume and volatility.


# Quick Start

Let's start by loading the package:To construct a reprex, I figured I better install keras. After installing keras, I can successfully call a tensorflow function, but not access anything in keras. The error says the lazy-load database is corrupt. Is that an obscure way to say my operating system (Ubuntu 14.04 LTS) is too old? Rumor has it that keras requires Ubuntu >= 16.04, though I haven't see that dependency stated explicitly, and it istalled without error.
```{r, echo=TRUE}
devtools::load_all()
```


# References {-}


