---
title: "Intraday Market Signal: Modeling and Prediction "
author: |
  | Shengjie Xiu, Yifan Yu, and Daniel P. Palomar
  | The Hong Kong University of Science and Technology (HKUST)
date: "`r Sys.Date()`"
output: 
  cleanrmd::html_document_clean:
    theme: "bamboo"
    mathjax: default
    toc: true
    toc_depth: 2
csl: apalike.csl # springer-socpsych-author-date.csl
csl-entry-spacing: 1.2em
bibliography: refs.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Intraday Volume and Volatility: Modeling and Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteKeyword{State-space Model}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.retina = 2,
  out.width = "85%",
  dpi = 96,
  pngquant = "--speed=1"
)
knit_hooks$set(pngquant = hook_pngquant)  # brew install pngquant
options(width=100)
```

```{css, echo = FALSE}
/* ensure cleanrmd is centered */
body {
  margin: 0 auto;
  max-width: 750px;
  padding: 2rem;
}

/* math is smaller */
.math {
  font-size: small;
}
```

------------------------------------------------------------------------

> This vignette illustrates the usage of the package `intradayModel`. The package applies state-space models to model and predict intraday market signal, specifically intraday trading volume. The future version of this package will support the intraday volatility.

# Quick Start

Let's load our package and the sample data: the 15-min intraday volume of AAPL from 2019-01-02 to 2019-06-28, including 124 trading days.

```{r, message = FALSE}
devtools::load_all()
data(AAPL_volume)
```

We define a univariate state-space model by the `uniModelSpec` method.

```{r}
model <- uniModelSpec(fit = TRUE)
```

Let's use the first 104 trading days to fit the model, and the last 20 days to evaluate the prediction performance. The model can be fitted to the training data using its `uniModelFit` method.

```{r}
data <- AAPL_volume
data_train <- AAPL_volume[, 1:104]
model_fitted <- uniModelFit(data_train, model, acceleration = TRUE)
```

Once the model is fitted, it can be used to decompose intraday trading signal into daily, seasonal, and intraday dynamic components via `uniModelFilter` method.

```{r}
filter_result <- uniModelFilter(data_train, model_fitted)
filter_result$plot
```

With the fitted model, we can do one-bin-ahead prediction on new data using `uniModelPred`. In this case, we predict on the last 20 trading days of the dataset, and evaluate its accuracy.

```{r}
predict_result <- uniModelPred(data, model_fitted, out.sample = 20)
predict_result$measure
predict_result$plot
```

&nbsp; 

# Usage of the package

## Preliminary knowledge

We denote the day with index $t\in\{1,\dots,T\}$ and divide each day into $I$ intervals (referred to as bins) indexed by $i\in\{1,\dots,I\}$ The intraday observations of market signal is labeled with the single subscript $\tau = I \times (t-1) + i$ to represent the index of time series.

The observed market signal can be written in the following multiplicative form [@brownlees2011intra]:

$$
\large
\text{intraday signal} = \text{daily} \times \text{seasonal} \times \text{intraday dynamic} \times \text{noise}. \tag{1}
\small
$$

By taking the logarithm transform, we rewrite (1) as an addictive formula:

$$
\large
y_{\tau} = \eta_{\tau} + \phi_i + \mu_{t,i} + v_{t,i}, \tag{2}
\small
$$

where each log-component can be extracted by the state-space model.

The univariate model proposed by [@chen2016forecasting] is defined as
$$
\large
\begin{aligned}
\mathbf{x}_{\tau+1} &= \mathbf{A}_{\tau}\mathbf{x}_{\tau} + \mathbf{w}_{\tau},\\
y_{\tau} &= \mathbf{C}\mathbf{x}_{\tau} + \phi_{\tau} + v_\tau,
\end{aligned}
\tag{3}
\small
$$
where:

- $\mathbf{x}_{\tau} = [\eta_{\tau}, \mu_{\tau}]^\top$ is the hidden state vector containing the daily component and the intraday component;

- $\mathbf{A}_{\tau} = \left[\begin{array}{l}a_t^{\eta}&0\\0&a^{\mu}\end{array} \right]$ is the state transition matrix with $a_t^{\eta} = \begin{cases}a^{\eta}&\tau = kI, k = 1,2,\dots\\0&\text{otherwise};\end{cases}$

- $\mathbf{C} = [1, 1]$ is the observation matrix;

- $\boldsymbol{\phi} = [\phi_1,\dots, \phi_I]^\top$ is the seasonal component;

- $\mathbf{w}_{\tau} = \left[\epsilon_t^{\eta},\epsilon_t^{\mu}\right]^\top \sim \mathcal{N}(\mathbf{0}, \mathbf{Q}_{\tau})$ represents the i.i.d. Gaussian noise in the state transition, with diagonal covariance matrix $\mathbf{Q}_{\tau} = \left[\begin{array}{l}(\sigma_\tau^{\eta})^2&0\\0&(\sigma^{\mu})^2\end{array} \right]$ and $\sigma_\tau^{\eta} = \begin{cases}\sigma^{\eta}&\tau = kI, k = 1,2,\dots\\0&\text{otherwise};\end{cases}$

- $v_\tau \sim \mathcal{N}(0, r)$ is the i.i.d. Gaussian noise in the observation;

- $\mathbf{x}_1$ is the initial state, and it is assumed to follow $\mathcal{N}(\mathbf{x}_0, \mathbf{V}_0)$.

In this model, $\boldsymbol{\Theta} = \{\mathbf{x}_0, \mathbf{V}_0, a^{\eta}, a^{\mu}, (\sigma^{\eta})^2, (\sigma^{\mu})^2, r, \boldsymbol{\phi} \}$ are treated as parameters.

## Datasets

To use this package, the time series object without missing values should be formatted into an $I \times T$, i.e., `(n_bin, n_day)`, 2D numeric matrix. For convenience, the package contains two sample datasets: `AAPL_volume` and `GE_volume`, including 15-min trading volume of AAPL and GE within 124 trading days from 2019-01-02 to 2019-06-28.

```{r}
data(GE_volume)
dim(GE_volume)
```

## Model specification

> **uniModelSpec**(fit = FALSE, fixed.pars = NULL, init.pars = NULL)

This function specifies a univariate state-space model. To set `fit=TRUE` if the model contains unknown parameters. If you want to fix parameters to some values, you can provide a list of values to `fixed.pars`. If you have prior knowledge of the initial values for the unfitted parameters, you can provide it through `init.pars`.

Here is a simple example.

```{r}
# set fixed value
fixed.pars <- list()
fixed.pars$"x0" <- c(13.33, -0.37)

# set initial value 
init.pars <- list()
init.pars$"a_eta" <- 1

# define the univariate model
model <- uniModelSpec(fit = TRUE, fixed.pars = fixed.pars, init.pars = init.pars)
```


## Fitting

> **uniModelFit**(data, uniModel, maxit = 3000, abstol = 1e-4, log.switch = TRUE, acceleration = FALSE, verbose = 1)

Once the model with unfitted parameters is defined, you can apply `uniModelFit` function to fit the model. The algorithm used is the expectation–maximization (EM) given in [@chen2016forecasting]. It terminates when iteration reaches `maxit` or the change of the parameters achieves $\|\Delta \boldsymbol{\Theta}_i\| \le \text{abstol}$.

If `log.switch = TRUE`, the intermediate parameters are recored during the algorithm. 

If `acceleration=TRUE`, the accelerated EM algorithm  [@varadhan2008simple] is applied with possibly lower numerical stability.

The `verbose` controls how much output is shown during the fitting process. Possible values:

- `0`: no output

- `1`: show iteration number and change of the parameters

- `2`: 1 + show obtained parameters

The following code shows how to fit the model to the GE stock.

```{r}
data <- GE_volume
data_train <- GE_volume[, 1:104]
model_fitted <- uniModelFit(data_train, model, acceleration = TRUE, verbose = 2)
```

## Filtering

> **uniModelFilter**(data, uniModel)

A model with all parameters fixed can be used to decompose the intraday trading signal into daily, seasonal, and intraday dynamic components. The daily component and intraday dynamic component at time $t$ are the smoothed state estimate conditioned on all the data. They are mathematically denoted by $\mathbb{E}[\mathbf{x}_{\tau}|\{y_{\tau}\}_{\tau=1}^{N}]$, where $N$ is the total number of bins in the train set. The seasonal component has the value of $\boldsymbol{\phi}$.

Input the intraday market signal that need to be decomposed and the fitted model into function `uniModelFilter`. This function will produce the three components along with a plot.

```{r}
filter_result <- uniModelFilter(data_train, model_fitted)
str(filter_result$components)
filter_result$plot
```

## Prediction

> **uniModelPred**(data, uniModel, out.sample) 

Once the model is fitted, you can predict the intraday market signal one-bin-ahead with the function `uniModelPred`. The one-bin-ahead prediction is mathematically denoted by $\hat{y}_{\tau+1} = \mathbb{E}[y_{\tau+1}|\{y_{j}\}_{j=1}^{\tau}]$.
Given the dataset, you need to indicate how many days from the end to keep for out-of-sample prediction.

Since fitting a model to some data does not entail that it will predict well on unseen data, `uniModelPred` also evaluates the model with the following measures for you.

- Mean absolute error (MAE): $\frac{1}{M}\sum_{\tau=1}^M\lvert\hat{y}_\tau - y_\tau\rvert$,

- Mean absolute percent error (MAPE): $\frac{1}{M}\sum_{\tau=1}^M\frac{\lvert\hat{y}_\tau - \sigma_\tau\rvert}{y_\tau}$,

- Root Mean Square Error (RMSE): $\sqrt{\sum_{\tau=1}^M\frac{\left(\hat{y}_\tau - y_\tau\right)^2}{M}}$,

where $M$ is the total number of out-of-sample bins.

```{r}
predict_result <- uniModelPred(data, model_fitted, out.sample = 20)
head(predict_result$signal_pred)
predict_result$measure
predict_result$plot
```

&nbsp; 

# References {.unnumbered}

