---
title: "Intraday Volume and Volatility: Modeling and Prediction"
author: |
  | Shengjie Xiu, Yifan Yu, and Daniel P. Palomar
  | The Hong Kong University of Science and Technology (HKUST)
date: "`r Sys.Date()`"
output: 
  cleanrmd::html_document_clean:
    theme: bamboo
    toc: yes
    toc_depth: 2
csl: apalike.csl # springer-socpsych-author-date.csl
csl-entry-spacing: 1.2em
bibliography: refs.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Intraday Volume and Volatility: Modeling and Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteKeyword{State-space Model}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.retina = 2,
  out.width = "85%",
  dpi = 96,
  pngquant = "--speed=1"
)
knit_hooks$set(pngquant = hook_pngquant)  # brew install pngquant
options(width=100)
```

```{css, echo = FALSE}
/* ensure cleanrmd is centered */
body {
  margin: 0 auto;
  max-width: 750px;
  padding: 2rem;
}
```

-----------
> This vignette illustrates the usage of the package `intradayModel` for modeling and prediction of intraday trading volume and volatility.

# Quick Start

Let's load our package and the sample data: the 15-min intraday volume of AAPL from 2019-01-02 to 2019-06-28, including 124 trading days. Each column contains 26 records of intraday volume at a trading day.

```{r}
devtools::load_all() # why here cannot use library(intradayModel)
data(AAPL_volume)
dim(AAPL_volume)
```

We define a univariate state-space model for the intraday signal by

```{r}
modelSpec <- uniModelSpec(fit = TRUE)
```

Let's use the first 104 trading days to fit the model, and the last 20 days to evaluate the prediction performance. To fit the model using the accelerated EM algorithm, we call

```{r}
data <- AAPL_volume
data_train <- AAPL_volume[, 1:104]
modelSpec_fitted <- uniModelFit(data_train, modelSpec, acceleration = TRUE)
```

To evaluate the fitting performance, we use the fitted model to decompose the intraday volume:

```{r}
filter_result <- uniModelFilter(data_train, modelSpec_fitted)
filter_result$plot
```

With the fitted model, we can do one-bin-ahead prediction and evaluate its accuracy.

```{r}
predict_result <- uniModelPred(data, modelSpec_fitted, out.sample = 20)
predict_result$measure
predict_result$plot
```

<!-- # Details (refer to rugarch vignette) -->


# References {-}


